---
title: "Informe IAM – Análisis clínico reproducible y auditable"
author: "Juan Munera"
date: "2026-01-29"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
# Global options
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Cardiovascular disease remains one of the leading causes of morbidity and mortality worldwide.

The first objective is to identify which variable are associated with the presence of cardiovascular disease in the dataset, through the variable "Heart Disease". Only information available at the initial clinical assessment is considered, in order to avoid information leakage and temporal bias.

The second one focuses on prognosis and clinical evolution among patients who have experienced acute myocardial infarction (IAM). Within this subgroup, we evaluate the association between selected clinical variables and short- and medium-term outcomes, such as recovery at discharge or prognosis at three months, considering the temporal availability of each variable. 

All analyses are interpreted in terms of association or prognostic value, without making causal claims.


## Data to be analyzed

For this analysis, we use a clinical data set of patients evaluated for cardiovascular disease, provided in Excel format.
The dataset contains information on 320 patients and 32 variables, including demographic characteristics, clinical measurements, biomarkets, treatment and follow-up outcomes.
In addition to the clinical dataset, a variable dictionary is provided. This one specifies the intended meaning, coding and when available, the units of each variable.


```{r Step 1}
#=============================================================
# Step 1: Load input data in a fully reproducible way
# No manual editing is allowed 

#Load required libraries

library(readxl)
library(janitor) # to clean and standardize column names

# Define file paths

dataPath <- "data/raw/Base_de_datos_IAM.xlsx"
dictPat <- "data/raw/Diccionario_variables.xlsx"

rawData<-read_excel(dataPath)
dictVars<-read_excel(dictPat)

# Clean column names to make them consistent 
# Example: "Heart Disease" -> "heart_disease"
rawData<-clean_names(rawData)
dictVars<-clean_names(dictVars)

dim(rawData)
names(rawData)

cat("\n") # Blank line to visually separate outputs

dim(dictVars)
names(dictVars)
```

## data audit purpose

The purpose of this step is to perform an initial audit of the variables contains in dataset, in order to understand its structure, types and basic internal coherence, as well as to compare the content of the Excel file with the variable dictionary.

At this stage, no data cleaning or modification is performed.The goal is strictly to identify potential issues such as incorrect data types, undocumented variables, discrepancies between the dataset and the dictionary, or values that may require further justification. These issues will be explicitly addressed and resolved in subsequent preprocessing steps.


```{r initial-audit, message=FALSE, warning=FALSE}
#=============================================================
# Step 2: Initial data audit 
# purpose: inspect structure, variable types

library(dplyr)
library(skimr)



# Store dataset dimensions in variables
dataDimensions<-dim(rawData)
nPatients<-dataDimensions[1]
nVariables<-dataDimensions[2]

# Print a clear message (useful in knitted reports)
cat("Dataset dimensions:", dataDimensions, 
    " patients:", nPatients,
    "and variables:", nVariables)

# Inspect structure and variable types
glimpse(rawData)
cat("\n")

#Global summary
#skimr::skim(rawData) # skim() provides per-variable summaries: type, missingness, min/mas, etc
cat("\n")

# Dataset against dictionary comparison
# Dictionary column containing variable name is : "variable
# Extract variable names from dictionary and dataset
dictVarName<-dictVars$variable |>as.character()|>na.omit()|>unique()
dataVarNames<-names(rawData)

# Variables documented in the dictionary but missing in dataset
varsInDict<- setdiff(dictVarName,dataVarNames)
varsInDataset<- setdiff(dataVarNames,dictVarName)
cat("Variables in dictionary but NOT in dataset: \n")
print(varsInDict)
cat("\nVariables in dataset but NOT in dictionary: \n")
print(varsInDataset)


```
## Plausibility checks

In this step we perform basic plausibility checks to identify values that are likely impossible or inconsistent (negative values where not expected, extreme out of range measurements), potential sentinel codes used to represent missingness, and temporal inconsistencies if date variables are available.

```{r plausibility-checks, message=FALSE, warning=FALSE}
#=============================================================
# Step 3: Plausibility checks
# - Numeric ranges (max/min) and extreme values
# - Potential sentinel values used to as missing codes
# - Simple temporal consistency checks (if date variables exist)

library(tidyr)

# Numeric ranges + missingness

# Identify numeric variables
numVars<- names(rawData)[sapply(rawData,is.numeric)]
# Summarize numeric variables
numRangeSummary<-rawData %>%
  summarise(
    across(
      all_of(numVars),
      list(
        n_missing= ~sum(is.na(.x)),
        pct_missing= ~mean(is.na(.x))*100,
        max= ~max(.x,na.rm=TRUE),
        min= ~min(.x,na.rm=TRUE)
      ),
      .names = "{.col}__{.fn}"))%>%
  pivot_longer(cols = everything(),
               names_to = c("variable","metric"),
               names_sep = "__",
               values_to = "value")%>%
  pivot_wider(names_from = metric,values_from = value)%>%
  arrange(desc(pct_missing)
  )

cat("Numeric variables range summary (sorted by % missing): \n")
print(numRangeSummary)

```

## Preprocesing and construction of the analytical datasets

In this step we construct two analysis datasets following a strict temporal logic to reduce bias and prevent information leakage.

Dataset A (Goal 1) includes only baseline variables and outcome.
Dataset B (Goal 2) is restricted to patients meeting the operational definition of IAM and includes variables that are temporally valid for prognosis.

```{r build-analytical-datasets, message=FALSE, warning=FALSE}
#=============================================================
# Step 4: Preprocesing + analytical dataset construction 
# - Basic typing helpers 
# - Create a temporal map using name heuristics
# - Build dataset goal 1 and dataset goal 2
# - Save dataset as .rds
library(stringr)
library(readr)

dictVars<-dictVars %>%
  mutate(
    variable=make_clean_names(variable))
    
# Basic preprocesing 
dataWork<-rawData

# Define the key variables 
requieredOutcome1<-"heart_disease"
requieredOutcome2<-"recovery_discharge"

if (
  requieredOutcome1 %in% names(dataWork)
){
  cat("Outcome variable found in dataset:", requieredOutcome1)
}else{
  
  stop("Outcome variable for Objective 1 is not found: heart_disease.")
}

# Continue workflow ONLY if the outcome exists

if (
    requieredOutcome2 %in% names(dataWork)
){
  cat("Outcome variable found in dataset:", requieredOutcome2)
}else{
  
  stop("Outcome variable for Objective 1 is not found: recovery_discharge")
}

# Audit: check distributions of outcomes
print(table(dataWork[[requieredOutcome1]], useNA = "ifany"))
print(table(dataWork[[requieredOutcome2]], useNA = "ifany"))

# Select baselines variables ("antes_del_evento")
baselineVars<-dictVars %>%
  filter(antes_despues_del_evento == "Antes") %>%
  pull(variable) %>%
  as.character()

# Keep only variables that actually exist in the dataset
baselineVars<- intersect(baselineVars,names(rawData))

# IMPORTANT: baseline predictors must not include the outcome
baselineVars<- setdiff(baselineVars,requieredOutcome1)

cat("Baseline variables (kwown before event: \n")
print(baselineVars)

# Build dataset A: outcome + baseline predictors
datasetA<-rawData %>%
  select(all_of(c(requieredOutcome1,baselineVars)))

# Build dataset B: IAM subject + chosen prognosis outcome + predictors
datasetB<-rawData %>%
  filter(.data[[requieredOutcome1]] == 1) %>%
  select(all_of(c(requieredOutcome2,baselineVars)))

# Save analytical datasets (.rds) and report dimensions
saveRDS(datasetA, "IAM_dataset_A.rds")
saveRDS(datasetB, "IAM_dataset_B.rds")

cat("\nDataset A dimensions:\n")
print(dim(datasetA))

cat("\nDataset B dimensions (IAM subset):\n")
print(dim(datasetB))

```

## Missing values (NA)

Missing data were systematically evaluated to assess data quality and guide the analytical strategy.
We quantified the proportion of missing values per variable, explored the missingness pattern, and compared two approaches: complete case analysis and simple imputation.Decisions were made considering potential bias, loss pf precision and interpretability.



```{r Mising values, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(naniar)

# Quantify missingness per variable (Dataseta A)

naSummaryA <- datasetA %>%
  summarise(
    across(
      everything(),
      list(
        nMissing = ~ sum(is.na(.x)), 
        pctMissing = ~ mean(is.na(.x))*100 
      ),
      .names ="{.col}__{.fn}"
    )
  ) %>%
  pivot_longer(
    cols= everything(),
    names_to = c("variable",  "metric"),
    names_sep = "__",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  arrange(desc(pctMissing))

print(naSummaryA)

# Strategy definition
# Exclude variables with excessive missingness (>40%)
varsKeep <- naSummaryA %>%
  filter(pctMissing <=40) %>%
  pull(variable)

datasetAReduced <- datasetA %>%
  select(all_of(varsKeep))
print(datasetAReduced)

# Approach: complete case analysis
datasetACc <- datasetAReduced %>%
  drop_na()

cat(" Complete case dataset dimensions: \n")
print(dim(datasetACc))
print(datasetACc)

# Approach: simple imputation
# Numeric: mean, Categorical: mode
getMode <- function(x){
   ux <- na.omit(unique(x))
   ux[which.max(tabulate(match(x,ux)))]
 }

datasetAImp <- datasetAReduced %>%
  mutate(
    across(
      where(is.numeric),
      ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)
    ),
    across(
      where(is.factor),
      ~ ifelse(is.na(.x), get_mode(.x), .x)
    )
  )

cat("Imputed dataset dimensions:\n")
print(dim(datasetAImp))
print(datasetAImp)
 
```

# Outliers and extrema values

Extreme values were assessed in numeric baseline variables to evaluate data quality and determine whether unusually large/small observations could unduly influence subsequent analyses. Outliers were identified visually (boxplots/histograms) and using an IQR-based rule. Rather than removing observations, a conservative approach was applied and a brief sensitivity analysis was conducted using winsorization to assess robustness.


```{r Outliers, warning=FALSE, message=FALSE}

# Identify numeric variables
numVarsA <- names(datasetAReduced)[sapply(datasetAReduced, is.numeric)]

cat("Numeric variables available in datset A reduced: \n")
print(numVarsA)

# Choose at least two numeric variables for outliers analysis 
if(length(numVarsA) <2) stop("Not enough numeric variables to perform outlier analyisis")

var1 <- numVarsA[1]
var2 <- numVarsA[2]

cat("\nVariables selected for outlier analysis:\n")
print(c(var1,var2))

# Visual inspection outliers
par(mfrow=c(2,2))

boxplot(datasetAReduced[[var1]],
        main = paste("Boxplot:", var1),
        ylab = var1)

hist(datasetAReduced[[var1]],
     main = paste("Histogram:", var1),
     xlab = var1)

boxplot(datasetAReduced[[var2]],
        main = paste("Boxplot:", var2),
        ylab = var2)

hist(datasetAReduced[[var2]],
     main = paste("Histogram:",var2),
     xlab = var2)


```
## Justified decision

The variables ekg_results and chest_pain_type are categorical variables encoded numerically. The extreme values identified through boxplots correspond to valid clinical categories rather than measurement errors. Therefore, the decision was to retain all observations without transformation, winsorization, or exclusion, and to treat these variables as categorical predictors in subsequent analyses.

## Sensitivity analysis

A sensitivity analysis based on outlier treatment was not performed, as the detected extreme values are intrinsic to the categorical coding of the variables. Consequently, outlier handling does not affect the main results, since no true anomalous values are present.

## EDA

```{r EDA, warning=FALSE, message=FALSE}

# Define outcome and check availability
y <- requieredOutcome1
if (!(y %in% names(datasetAReduced))){
  stop("Outcome variable not found in dataset reduced")
}

# Ensure outcome is categorical fot stratification 
datasetAReduced[[y]]<-as.factor(datasetAReduced[[y]])

# Predictors = all variables except the outcome
vars <- setdiff(names(datasetAReduced),y)

# Summary for table 1

summarizeNumeric <- function(x){
  c(
    nObs= sum(!is.na(x)),
    mean= mean(x,na.rm=TRUE),
    sd= sd(x,na.rm=TRUE),
    median=median(x,na.rm=TRUE),
    q1= as.numeric(quantile(x,0.25,na.rm=TRUE)),
    q3= as.numeric(quantile(x,0.75,na.rm=TRUE))
  )
}

summarizeCategorical <- function(x){
  tab<- table(x,useNA = "ifany")
  prop <-prop.table(tab)
  data.frame(
    level=names(tab),
    n= as.integer(tab),
    pct=round(as.numeric(prop)*100,1),
    row.names = NULL
  )
}

# Build table 1 objects
table1Overall <-list()
table1ByOutcome <-list()

for (variable in vars) {
  x<-datasetAReduced[[variable]]
  
  if(is.numeric(x)){
    table1Overall[[variable]] <-summarizeNumeric(x)
    table1ByOutcome[[variable]] <- lapply(split(x,datasetAReduced[[y]]),summarizeNumeric)
  }else{
    table1Overall[[variable]] <-summarizeCategorical(x)
    table1ByOutcome[[variable]] <- lapply(split(x,datasetAReduced[[y]]),summarizeCategorical)
  }
}

cat("Table 1 created as objects: table1_overall and table1_by_outcome\n")

# Print some variables
cat("\nExample - overall summary for first two variables:\n")
print(table1Overall[[vars[1]]])
print(table1Overall[[vars[2]]])

cat("\nExample - stratified summary for the first variable:\n")
print(table1ByOutcome[[vars[1]]])

# Figure about distribution of a key numeric variable
numKey <- vars[sapply(datasetAReduced[vars],is.numeric)]
if(length(numKey) ==0) stop("Variables are not avaible for EDA figures")

keyVar <- numKey[1]

hist(
  datasetAReduced[[keyVar]],
  main= paste("Distribution of", keyVar),
  xlab= keyVar
)

```
